---
title: "Portfolio Optimization Practice"
author: "Alex Kong"
date: "2/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(portfolioBacktest)
```
# R session: Loading market data

```{r}
library(xts)  # to manipulate time series of stock data
library(quantmod)  # to download stock data
library(PerformanceAnalytics)  # to compute performance measures
library(CVXR)

# download data from YahooFinance
stock_namelist <- c("AMD", "ADI", "APD", "BA", "DD", "M", "JNPR", "INTC", "GS", "IBM", "NOK", "XOM", "WFC")
prices <- xts()
for (i in 1:length(stock_namelist)) {
  tmp <- Ad(getSymbols(stock_namelist[i], from = "2005-01-01", to = "2008-12-31", auto.assign = FALSE))
  tmp <- na.approx(tmp, na.rm = FALSE)  # interpolate NAs
 # tmp <- na.approx(tmp, na.rm = TRUE)  # interpolate NAs
  prices <- cbind(prices, tmp)
}
colnames(prices) <- stock_namelist
tclass(prices) <- "Date"

head(prices)
tail(prices)

# compute log-returns and linear returns
X_log <- diff(log(prices))[-1]
X_lin <- (prices/lag(prices) - 1)[-1]
# or alternatively...
X_log <- CalculateReturns(prices, "log")[-1]
X_lin <- CalculateReturns(prices)[-1]

head(X_lin)

N <- ncol(X_log)  # number of stocks
T <- nrow(X_log)  # number of days
cat("N:", N, "T:", T, "\n")
```


Take a look at the share price

```{r}
plot(prices/rep(prices[1, ], each = nrow(prices)), col = rainbow10equal, legend.loc = "topleft",
     main = "Normalized prices")
```

Divide data into a training set and a test set

```{r}
T_trn <- round(0.7*T)  # 70% of data
X_log_trn <- X_log[1:T_trn, ]
X_log_tst <- X_log[(T_trn+1):T, ]
X_lin_trn <- X_lin[1:T_trn, ]
X_lin_tst <- X_lin[(T_trn+1):T, ]
```

Estimate mu and sigma with the training set
```{r}
#mu <- colMeans(X_log_trn)
mu <- colMeans(X_log_trn, na.rm = TRUE)
#Sigma <- cov(X_log_trn)
Sigma <- cov(X_log_trn)
mu
Sigma
```

Buy & Hold Strategy
```{r}
# a B&H portfolio is trivially the zero vector with a one on the stock held
w_BnH <- diag(N)
rownames(w_BnH) <- colnames(X_lin)
colnames(w_BnH) <- paste0("B&H-", colnames(X_lin))
dim(w_BnH)
dim(X_lin)
w_BnH


# compute returns of all B&H portfolios
ret_BnH <- xts(X_lin %*% w_BnH, index(X_lin))
ret_BnH_trn <- ret_BnH[1:T_trn, ]
ret_BnH_tst <- ret_BnH[-c(1:T_trn), ]
head(ret_BnH)
tail(ret_BnH, n = 10)

# performance measures
t(table.AnnualizedReturns(ret_BnH_trn))
t(table.AnnualizedReturns(ret_BnH_tst))

table.DownsideRisk(ret_BnH_trn)
```

To compute the wealth or cumulative P&L, we have two options: one assumes the
same quantity is repeateadly invested, whereas the other assumes
reinvesting (compounding):

```{r}
# compute cumulative wealth
dim(ret_BnH_trn)
wealth_arith_BnH_trn <- 1 + cumsum(ret_BnH_trn)  # initial budget of 1$
wealth_geom_BnH_trn <- cumprod(1 + ret_BnH_trn)  # initial budget of 1$

# plots
# same as: 
#   plot(wealth_arith_BnH_trn[, 1], main = "Buy & Hold performance (not compounded)", ylab = "wealth")
chart.CumReturns(ret_BnH_trn[, 1], main = "Buy & Hold performance (not compounded)", 
                 geometric = FALSE, wealth.index = TRUE)

# same as: 
#   plot(wealth_geom_BnH_trn[, 1], main = "Buy & Hold performance (compounded)", ylab = "wealth")
chart.CumReturns(ret_BnH_trn[, 1], main = "Buy & Hold performance (compounded)", 
                 geometric = TRUE, wealth.index = TRUE)

# more plots
chart.CumReturns(ret_BnH, main = "Buy & Hold performance", 
                 wealth.index = TRUE, legend.loc = "topleft", colorset = rich10equal)

charts.PerformanceSummary(ret_BnH_trn, main = "Buy & Hold performance", 
                          wealth.index = TRUE, colorset = rich10equal)

chart.Boxplot(ret_BnH_trn)

chart.RiskReturnScatter(ret_BnH_trn, symbolset = 21, bg = "red")
```

# Comparison of $\frac{1}{N}$ portfolio, quintile portfolio, and GMRP

The EWP or $\frac{1}{N}$ portfolio allocates equal dollar weight to each stock: 
**w**$=\frac{1}{N}$__1__.

```{r}
w_EWP <- rep(1/N, N)
names(w_EWP) <- colnames(X_lin)
w_EWP
```

__Quintile portfolios__ are widely used by practitioners. The idea is to 1) rank the N stocks, 2) divide them into five parts, and 3) long the top part (and possibly short the bottom part). One can rank the stocks in a multitude of ways (typically based on expensive factors that investment funds buy at a premium price). For our experiments, we will consider three possible rankings according to:

1. $\mu$
2. $\frac{\mu}{diag(\Sigma)}$
3. $\frac{\mu}{diag(\sqrt{\Sigma})}$

```{r}
# find indices of sorted stocks
i1 <- sort(mu, decreasing = TRUE, index.return = TRUE)$ix
i2 <- sort(mu/diag(Sigma), decreasing = TRUE, index.return = TRUE)$ix
i3 <- sort(mu/sqrt(diag(Sigma)), decreasing = TRUE, index.return = TRUE)$ix
mu
i1
i2
i3

# create portfolios
w_QuintP_1 <- w_QuintP_2 <- w_QuintP_3 <- rep(0, N)
w_QuintP_1[i1[1:round(N/5)]] <- 1/round(N/5)
w_QuintP_2[i2[1:round(N/5)]] <- 1/round(N/5)
w_QuintP_3[i3[1:round(N/5)]] <- 1/round(N/5)
w_QuintP <- cbind("QuintP (mu)"        = w_QuintP_1, 
                  "QuintP (mu/sigma2)" = w_QuintP_2, 
                  "QuintP (mu/sigma)"  = w_QuintP_3)
rownames(w_QuintP) <- colnames(X_lin)
w_QuintP
```

The __global maximum return portfolio (GMRP)__ chooses the stock with the hightest return during the in-sample period:

```{r}
i_max <- which.max(mu)
w_GMRP <- rep(0, N)
w_GMRP[i_max] <- 1
names(w_GMRP) <- colnames(X_lin)
w_GMRP
```

We can now compare the allocations of the portfolios:
```{r}
# put together all portfolios
w_heuristic <- cbind("EWP" = w_EWP, w_QuintP, "GMRP" = w_GMRP)
round(w_heuristic, digits = 2)

barplot(t(w_heuristic), col = rainbow8equal[1:5], 
        legend = colnames(w_heuristic), beside = TRUE,
        main = "Portfolio allocation of heuristic portfolios", 
        xlab = "stocks", ylab = "dollars")
```

Then we can compare the performance (in-sample vs out-of-sample):

```{r}
# compute returns of all portfolios
ret_heuristic <- xts(X_lin %*% w_heuristic, index(X_lin))
#ret_heuristic$`QuintP (mu/sigma2)` <- NULL  # remove since it coincides with "QuintP (mu/sigma)"
ret_heuristic_trn <- ret_heuristic[1:T_trn, ]
ret_heuristic_tst <- ret_heuristic[-c(1:T_trn), ]

# performance
t(table.AnnualizedReturns(ret_heuristic_trn))
t(table.AnnualizedReturns(ret_heuristic_tst))
```

Let’s plot the wealth evolution (cumulative PnL) over time:

```{r}
chart.CumReturns(R = ret_heuristic, main = "Cumulative return of heuristic portfolios", 
                 wealth.index = TRUE, legend.loc = "topleft", colorset = rich8equal)
addEventLines(xts("training", index(X_lin[T_trn])), srt = 90, pos = 2,
              lwd = 2, col = "darkblue")
```

Finally, we can plot the risk-return scatter plot:

```{r}
chart.RiskReturnScatter(R = ret_heuristic_trn, symbolset = 21, bg = "red",
                        main = "Annualized Return and Risk (in-sample)")
```
# Markowitz's mean-variance portfolio (1952)

Markowitz's mean-variance portfolio (MVP) with no shorting is formulated as:
$$
\underset{w}{maximize}\ \  \mu^T w- \lambda w^T \Sigma w \\
\text{subject to  } 1^T w = 1 \\
w \ge 0
$$
This problem does not have a closed-form solution so we use a solver.
```{r}
# create function for MVP
MVP <- function(mu, Sigma, lmd = 0.5) {
  w <- Variable(nrow(Sigma))
  prob <- Problem(Maximize(t(mu) %*% w - lmd*quad_form(w, Sigma)),
                  constraints = list(w >= 0, sum(w) == 1))
  result <- solve(prob)
  w <- as.vector(result$getValue(w))
  names(w) <- colnames(Sigma)
  return(w)
}

# this function can now be used as
w_MVP <- MVP(mu, Sigma, lmd = 2)
w_MVP

w_heuristic <- cbind("EWP" = w_EWP, w_QuintP, "GMRP" = w_GMRP, "MVP" = w_MVP)
round(w_heuristic, digits = 2)
barplot(t(w_heuristic), col = rainbow8equal[1:6], 
        legend = colnames(w_heuristic), beside = TRUE,
        main = "Portfolio allocation of heuristic portfolios", 
        xlab = "stocks", ylab = "dollars")
```

# Maximum Shapre ratio portfolio (MSRP)

* Markowitz’s mean-variance framework provides portfolios along the Pareto-optimal
frontier and the choice depends on the risk-aversion of the investor.

* But typically one measures an investment with the Sharpe ratio: only one
portfolio on the Pareto-optimal frontier achieves the maximum Sharpe ratio.

* Precisely, Sharpe (1966) first proposed the maximization of the Sharpe ratio:

$$
\underset{w}{maximize}\ \  \frac{w^T \mu - r_f}{\sqrt{w^T \Sigma w}} \\
\text{subject to  } 1^T w = 1 \\
w \ge 0
$$